---
title: "Multivariate Lake Data Example"
author: "Thomas Petzoldt"
date: "2021-05-12"
output: html_document
bibliography: bib.bib
---

```{r setup, include=FALSE}
## this code chunk sets some technical details, it is not shown to the user
library("readxl")
library("vegan")
#knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
#setupKnitr(autoprint = TRUE)
```

## Introduction

The following example demonstrates basic multivariate principles by means of a 
teaching example. A few more examples are available in  the lecture tutorial
[@Petzoldt2014elements]. The detailed theory and applications is found in 
excellent books of @Legendre1998 and @Borcard2018. Practical help is found in the
tutorials of the **vegan** package [@Oksanen2020].

## Data set and terms of use

The lake data set originates from the public data repository of the German 
Umweltbundesamt [@UBA2020]. The data set provided can be used freely according
to the [terms and conditions](https://www.umweltbundesamt.de/datenschutz-haftung#c-urheber-und-kennzeichenrecht) 
published at the [UBA web site](https://www.umweltbundesamt.de/), that refer to
ยง 12a EGovG with respect of the data, and to the [Creative Commons CC-BY ND International License 4.0 ](https://creativecommons.org/licenses/by-nd/4.0/deed.de)
with respect to other objects directly created by UBA.

The document and codes provided here can be shared according to [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/deed.en).

## Load the data

Here we load the data set and add English column names and 
abbreviated lake identifiers as row names to the table,
that are useful for the multivariate plotting functions.

```{r}
library("readxl") # read Excel files directly
library("vegan")  # multivariate statistics in ecology
lakes <- as.data.frame(
  read_excel("3_tab_kenndaten-ausgew-seen-d_2021-04-08.xlsx", sheet="Tabelle1", skip=3)
)
names(lakes) <- c("name", "state", "drainage", "population", "altitude", 
                  "z_mean", "z_max", "t_ret", "volume", "area", "shore_length", 
                  "shore_devel", "drain_ratio", "wfd_type")
rownames(lakes) <- paste0(1:nrow(lakes), substr(lakes$name, 1, 4))
```

Text columns, e.g Federal State names and lake type are removed and rows with missing data 
excluded. If population is not used, the analysis can be repeated with more lakes.

```{r}
valid_columns <- c("drainage", "population", "altitude", "z_mean",
                   "z_max", "t_ret", "volume", "area", "shore_length", 
                   "shore_devel", "drain_ratio")

#valid_columns <- c("drainage", "altitude", "z_mean",
#                   "z_max", "t_ret", "volume", "area", "shore_length", 
#                   "shore_devel","drain_ratio")
dat <- lakes[valid_columns]
dat <- na.omit(dat)
```

## Data inspection

It is alwas a good idea to plot the data first, as time series or boxplots for 
example, dependingon the type of data. Here we use boxplots, that we scale (z-transform) 
to a mean zero and standard deviation one to have comparable values.

As we can see a number of high extreme values, we apply also a square root 
transformation, that is less extreme than log transform and not sensitive against 
zero values, but because altitude contains a negative value (below sea level) we 
replace this with zero. As it is a small value, it does not influence our analysis,
but we should always be very careful to document such workarounds.

```{r}
par(mfrow=c(1,1))
par(mar=c(7,4,2,1)+.1)
boxplot(scale(dat), las=2)
dat$altitude <- ifelse(dat$altitude < 0, 0, dat$altitude)
boxplot(scale(sqrt(dat)), las=2)
```

## Multivariate Analysis

### Principal Components: PCA

```{r}
pc <- prcomp(scale(dat))
summary(pc)
plot(pc)
biplot(pc)
```

As the PCA with the untransformed data looks somewhat asymmetrix, we repeat it with
square transformed data. In addition, also the 3rd PC is plotted.

```{r}
dat2 <- sqrt(dat)
pc2 <- prcomp(scale(dat2))
summary(pc2)

par(mfrow=c(1,2))
par(mar=c(5, 4, 4, 2) + 0.1)
biplot(pc2, cex=0.6)
biplot(pc2, cex=0.6, choices=c(3,2))
```

A PCA is also possible with the `rda` function of the **vegan** package. The
syntax of the plot functions is somewhat different here. Instead of `biplot`as 
above, we can directly use `plot`. Details are foudn in the **vegan** documentation.

```{r}
par(mfrow=c(1,1))
pc3 <- rda(dat2, scale = TRUE)
pc3
#summary(pc3)
plot(pc3)
```

### Nonmetric Multidimensional Scaling: NMDS

And now an NMDS for the data set. Function `metaMDS` runs a series of NMDS runs
with different start values to avoid local minima. It has also some automatic
transformations built in and works usually with the Bray-Curtis dissimilarity,
that is used for plants and animal species abundance data. As we work with
physical data here, we set the distance measure to "euclidean".

```{r}
md <- metaMDS(dat2, scale=TRUE, distance = "euclid")
plot(md, type="text")
abline(h=0, col="grey", lty="dotted")
abline(v=0, col="grey", lty="dotted")
```

### Cluster analysis

Here we apply a hierarchical cluster analysis with square root transformed data 
and two different agglomeration schemes, "complete linkage" and "Ward's method".

```{r  fig.height=10}
par(mfrow=c(2,1))
hc <- hclust(dist(scale(dat2)), method="complete") # the default
plot(hc)

hc2 <- hclust(dist(scale(dat2)), method="ward.D2")
plot(hc2)
```

We can also use the clusters to indicate groups in the NMDS plot. Function 
`rect.hclust` indicates a given number of clusters in the dendrogram, then we cut
the tree with `cutree` and use the groups `grp` as color codes. **R** has 8 
standard colors. If we need more, we must define an own palette.

```{r}
plot(hc, hang = -1)
rect.hclust(hc, 5)
grp <- cutree(hc, 5)
# grp                  # can be used to show the groups
plot(md, type = "n")
text(md$points, row.names(dat2), col = grp)
```


Instead of a hierarchical clustering, we can also use a non-hierarchical method,
e.g. k-means clustering. This is an iterative method, and avoids the problem that 
cluster assignment depends on the order of clustering and the agglomeration method.

Depending on the question, it may be a disadvantage, that the number of clusters 
needs to be specified beforehand (e.g. from hierarchical clustering) and that we
do not get a tree diagramm.


## Task

* Try to understand the analysis, 
* discuss the results, 
* ask questions.
* The idea is to work on this report together and to make it more complete.

## References
